### 2013-05-31 doc/nips2013/factorial.tex

So far, we have relied on the existence of a bottleneck hidden variable $h_j$ on which
we have three conditionally independent views.  Such bottlenecks do not always
exist; for example, in \reffig{factorialModels}(a), neither $h_1$ or $h_2$ is a bottleneck.
Our solution is to create a composite latent variable $c$ which is in
one-to-one correspondence with the existing latent variables $(h_1,h_2)$.
We can invoke $\TensorFactorize$ to obtain composite marginals and
conditional means.  However, the identity of $h_1$ and $h_2$ becomes lost in the composite,
and must be recovered.  To do this, we develop a new algorithm $\UnshuffleFactorize$
to produce constraints on the original parameters.

\Fig{figures/factorialModels}{0.3}{factorialModels}{The factorial mixture model (a)
does not have three-view bottlenecks, so we first identify composite bottlenecks,
and then perform an \emph{unshuffling factorization} (b):
Given known composite conditional means $B_c = \BP[x_j \mid c] \in \R^\ell$,
which are derived from combinations of unknown parameters $f_a = \theta_1(\cdot, a), g_b = \theta_2(\cdot, b)$,
we group differences into bins (dashed boxes), and maximal bins into sources (red rounded boxes),
producing a linear program with variables $f_a,g_b,Z_{ab}$ and constraints for each $B_c$.
}

%restricted Boltzmann machines \cite{salakhutdinov09softmax}

We focus on the factorial mixture model.
Let $h = (h_1, \dots, h_s) \in [\nh]^s$ be the latent variables (sources)
and $x = (x_1, x_2, x_3) \in [\nphix]^3$ be the observed variables.
\begin{align}
p_\theta(x, h) = \prod_{i=1}^s p_\theta(h_i) \prod_{j=1}^3 p_\theta(x_j \mid h), \quad
p_\theta(x_j \mid h) = \exp\left\{ \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta; h) \right\},
\end{align}
where each $\theta_i \in \R^{\nphix \times \nh}$ are the parameters for the $i$-th source.
%Let $\theta = (\theta_1, \dots, \theta_b) \in \R^{\nphix \times b \nh}$.

@@

First, we run $\TensorFactorize$ on estimates of
$M_2 \eqdef \E[x_1 \otimes x_2]$ and $M_3 \eqdef \E[x_1 \otimes x_2 \otimes x_3]$,
which produces $B = (\E[x_j \mid c])_{c \in [\nh]^s}$.
%Next, we still have to extract the actual parameters $\theta$ from $B$. 
In the simple mixture model,
the clusters produced by $\TensorFactorize$ are in exact one-to-one correspondence
with the actual mixture components, but in the factorial case, we must solve
a \emph{credit assignment problem} to match combinations of $s$ sources
with columns of $B$.

Let $\{ v_{ia} \}_{i \in [s], a \in [k]}$, where $v_{ia} = \theta_i(\cdot, a) \in \R^\ell$,
be the collection of parameter vectors that we seek to recover.
Let $C \in \R^{\ell \times k^s}$ be the elementwise logarithm of $B$; that is
$C(x_j, h) = \log B(x_j, h)$.
The columns of $C$ are exactly the unordered collection of vectors
$\{ \sum_{i=1}^s v_{ia_i} - Z_{\ba} : \ba = (a_1, \dots, a_s) \in [\nh]^s \}$,
where $Z_{\ba} = \log \sum_{e=1}^\ell (\sum_{i=1}^s \exp(v_{ia_i}(e)))$ is the normalization constant (a scalar).

Note that there is a combinatorial dimension to this problem since the association between
columns of $C$ and combinations of $\{ v_{ia} \}$ is unknown.
As a running example (\reffig{factorialModels}(b)),
consider $s=2$ sources, each taking on $k=3$ values (let $f_a = v_{1a}$ and $g_b = v_{2b}$).
The following algorithm creates a linear program which puts constraints on variables $\{v_{ia}\}$ and $Z_{\ba}$:
\begin{enumerate}

\item Compute the difference vector $B_i - B_j$ for every distinct pair of columns $c,d \in [\nh^s]$.

\item Put the pairs $(c,d)$ into bins, where the difference vectors $B_c-B_d$
in each bin are \emph{comparable}, by which we mean that they
differ by a constant scalar times $\bone$ (for the normalization constant).

\item Keep only the maximal bins.
These bins should contain $\nh^{s-1}$ pairs, each corresponding to a fixed setting of the $s-1$ sources
and changing the value of one source.
Specifically, the difference vector is equal to $v_{ia} - v_{ia'}$ up to a
constant scalar for some source $i \in [k]$ and distinct values $a, a' \in
[\nh]$.

\item Group the bins into sources:
  For each source $i = 1, \dots, s$,
  chose any bin, declare it as $v_{i1} - v_{i2}$, and add it to the source.
  Let $\gamma$ be the set of $c$ such that that bin contains some difference vector $B_c - B_d$,
  and include any other bin with the same $\gamma$; these should correspond to $v_{i1} - v_{ia}$ for $a \ge 3$.
  Then throw away any bin whose difference vector is $-(B_c - B_d)$
  or is comparable to the difference of two existing difference vectors (e.g., $v_{i2} - v_{i1}$ or $v_{i2} - v_{i3}$).
  The resulting source should have $\nh-1$ bins, corresponding to $\{ v_{i1} - v_{ia} : 2 \le a \le \nh \}$.
  At this point, we have identified which vectors contribute to each $B_c$,
  taking care of the combinatorial aspect of the problem,
  but have yet to recover the actual vectors.

@@

  It turns out that this is LP is rank deficient for two reasons, the first intrinsic to our model,
  and the second due to loose linearization.
  First, the vectors $v_{ia}$ are non-identifiable since we can always add any vector $u \in \R^\ell$ to
  all the vectors of a source $i_1$ ($v_{i_1 a}$ for all $a \in [\nh]$)
  subtract $u$ from all the vectors of another source $v_{i_2 a}$ for all $a \in [\nh]$.
  Second, if we knew the normalization constants $\{Z_\ba\}$, then this is only rank deficiency.
  However, treating $\{Z_\ba\}$ as independent variables creates additional rank deficiency.

  Despite this, in the spirit of providing partial information,
  $\UnshuffleFactorize$ still gives us linear constraints on the parameters.
  In particular, we propose adding to the likelihood objective a strong regularization term
  $R(\theta, Z) = \frac{\lambda}{2} (\bA_1 \theta - \bA_2 Z - B)^2$, where $\bA_1$ and $\bA_2$ are coefficients
  recovered by unshuffling.
  In particular, we propose optimizing $\min_{\theta, Z} \sum_{i=1}^n -\log p_\theta(x^{(i)}) + R(\theta, Z)$.
  Although the resulting objective is still non-convex, the regularization imposes quite strong constraints,
  and the optimization would happen effectively over a much lower-dimensional subspace.
\end{enumerate}

>>>>

So far, we have relied on the existence of a bottleneck hidden variable $h_j$ on which
we have three conditionally independent views $x_{j_1},x_{j_2},x_{j_3}$.  Such bottlenecks do not always
exist; for example, in \reffig{factorialModels}(a), neither $h_1$ nor $h_2$ is
a bottleneck.
In this section, we present a technique to handle such models.
We start by creating a composite latent variable $c$ which is in
one-to-one correspondence with the existing latent variables $(h_1,h_2)$.
We can invoke $\TensorFactorize$ to obtain composite marginals and
conditional means, but the identities of $h_1$ and $h_2$ becomes lost in the
composite.  We develop a new algorithm $\UnshuffleFactorize$
to restore the identities and place constraints on the original
parameters.

\Fig{figures/factorialModels}{0.3}{factorialModels}{The factorial mixture model (a)
does not have three-view bottlenecks, so we first identify composite bottlenecks,
and then perform an \emph{unshuffling factorization} (b):
Given known composite conditional means
$L_c = \log \BP[x_j \mid c] \in \R^\ell$,
which are derived from combinations of unknown parameters
$f_a = \theta_1(\cdot, a), g_b = \theta_2(\cdot, b)$ and log-normalization scalars $Z_{ab}$,
we group difference vectors into bins (dashed boxes), and maximal bins into sources (red rounded boxes),
producing a linear program with variables $\{f_a\},\{g_b\},\{Z_{ab}\}$ and
a constraint for each $B_c$.
}

We focus on the restricted Boltzmann machine
\cite{smolensky86rbm,salakhutdinov09softmax},
which we refer to as an undirected factorial mixture model.
Let $h = (h_1, \dots, h_s) \in [\nh]^s$ be the latent variables (sources)
and $x = (x_1, x_2, x_3) \in [\nphix]^3$ be the observed variables.
The joint model probability is as follows:
\begin{align}
%p_\theta(x, h) = \prod_{i=1}^s p_\theta(h_i) \prod_{j=1}^3 p_\theta(x_j \mid h), \quad
%p_\theta(x_j \mid h) = \exp\left\{ \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta; h) \right\},
p_\theta(x, h) = \exp \left\{ \sum_{j=1}^3 \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta) \right\},
\end{align}
where each $\theta_i \in \R^{\nphix \times \nh}$ is the parameter matrix for the $i$-th source.

@@

First, we run $\TensorFactorize$ on estimates of
$M_2 \eqdef \E[x_1 \otimes x_2]$ and $M_3 \eqdef \E[x_1 \otimes x_2 \otimes x_3]$,
which produces $B = (\E[x_j \mid c])_{c \in [\nh]^s}$.
%Next, we still have to extract the actual parameters $\theta$ from $B$. 
In the simple mixture model,
the clusters produced by $\TensorFactorize$ are in exact one-to-one correspondence
with the actual mixture components, but in the factorial case, we must solve
a \emph{credit assignment problem} to match combinations of the $s$ sources
with columns of $B$.

Let $\{ v_{ia} \}_{i \in [s], a \in [k]}$, where $v_{ia} = \theta_i(\cdot, a) \in \R^\ell$,
be the collection of parameter vectors that we seek to recover.
Let $L \in \R^{\ell \times k^s}$ be the elementwise logarithm of $B$; that is
$L(x_j, h) = \log B(x_j, h)$.
The key observation is that the columns of $L$ are exactly the unordered collection of vectors
$\{ \sum_{i=1}^s v_{ia_i} - Z_{\ba} : \ba = (a_1, \dots, a_s) \in [\nh]^s \}$,
where $Z_{\ba} = \log \sum_{e=1}^\ell (\sum_{i=1}^s \exp(v_{ia_i}(e)))$ is the
log-normalization constant (a scalar) that sums over possible values $e$ of $x_j$.

Note that there is a combinatorial dimension to this problem since the association between
columns of $L$ and combinations of $\{ v_{ia} \}$ is unknown.
For example, example (\reffig{factorialModels}(b)),
consider $s=2$ sources, each taking on $k=3$ values
(for convenience, let $f_a = v_{1a}$ and $g_b = v_{2b}$).
The following algorithm, $\UnshuffleFactorize$,
creates a linear program that places constraints on
variables $\{v_{ia}\}$ and $Z_{\ba}$:
\begin{enumerate}

\item Compute the difference vector $L_c - L_d$ for every distinct pair of
columns $c,d \in [\nh^s]$.

\item Put the pairs $(c,d)$ into bins such that the difference vectors $L_c-L_d$
in each bin are \emph{comparable}, by which we mean that they
differ by a constant scalar times $\bone$ (for the normalization constant).

\item Keep only the maximal bins.
Such bins should contain $\nh^{s-1}$ pairs, each corresponding to a fixed
setting of the $s-1$ sources
and changing the value of only one source.
Specifically, the difference vector is equal to $v_{ia} - v_{ia'}$ up to a
constant scalar for some source $i \in [k]$ and distinct values $a, a' \in
[\nh]$.

\item Group the bins into sources:
  For each source $i = 1, \dots, s$,
  chose any bin $\omega$, declare it as $v_{i1} - v_{i2}$,
  and add $\omega$ to the source.
  Let $\gamma$ be the set of $c$ such that $\omega$
  contains some difference vector $L_c - L_d$,
  and include any other bin $\omega'$ with the same $\gamma$;
  $\omega$' should correspond to $v_{i1} - v_{ia}$ for $a \ge 3$.

  Then throw away any bin whose difference vector is $-(L_c - L_d)$
  or is comparable to the difference of two existing difference vectors
  (we want to discard bins corresponding to $v_{i2} - v_{i1}$ and $v_{i2} - v_{i3}$).
  Finally, the resulting source should have $\nh-1$ bins, corresponding to $\{ v_{i1} - v_{ia} : 2 \le a \le \nh \}$.
  At this point, we have identified which vectors contribute to each $L_c$,
  taking care of the combinatorial aspect of the problem,
  but have yet to recover the actual vectors.

@@

\end{enumerate}

It turns out that this is LP is rank deficient for two reasons,
the first intrinsic to our model,
and the second due to loose linearization.
First, the vectors $v_{ia}$ are non-identifiable since we can always add any vector $u \in \R^\ell$ to
all the vectors of a source $i_1$ ($v_{i_1 a}$ for all $a \in [\nh]$)
and subtract $u$ from all the vectors of another source $v_{i_2 a}$ for all $a \in [\nh]$.
Second, if we treated log-normalization constants $\{Z_\ba\}$ as a deterministic
function of $v_{ia}$, then the resulting program would have no other sources
of rank deficiency, but would also be non-linear.
Treating $\{Z_\ba\}$ as independent variables creates additional rank
deficiency.

Despite this, in the spirit of providing partial information,
we can still use the linear constraints provided by $\UnshuffleFactorize$.
In particular, define a regularization term
$R(\theta, Z) = \frac{1}{2} (\bA_1 \theta - \bA_2 Z - B)^2$, where $\bA_1$ and $\bA_2$ are coefficients
recovered by unshuffling.
In particular, we propose optimizing $\min_{\theta, Z} \sum_{i=1}^n -\log
p_\theta(x^{(i)}) + \lambda R(\theta, Z)$,
where $\lambda$ will be relatively large.
Although the resulting objective is still non-convex,
the regularization imposes quite strong constraints,
resulting in an optimization problem
that effectively operates over a much lower-dimensional subspace.

### 2013-05-31 doc/nips2013/mixture.tex

\Fig{figures/simpleModels}{0.3}{simpleModels}{Three-view mixture models: conditioned on a latent variable $h \in [\nh]$,
we have three conditionally independent observed variables.  We consider
estimation for both directed (left) and undirected (right) versions.
}

In this section, we consider the simplest model to illustrate the form of our
method.  Consider the undirected mixture model model in \reffig{simpleModels}(b).
We have one discrete latent variable $h \in [\nh]$ and
three observed variables
$x = (x_1, x_2, x_3) \in \sX_1 \times \sX_2 \times \sX_3 = \sX$
which are conditionally independent given $h$.
Formally, the model is as follows:\footnote{We assume for simplicity that the $x_j$'s are identically distributed conditioned on $h$,
but it is possible to use different distributions (see \citet{anandkumar12moments}).}
\begin{align}
  \label{eqn:mixtureModel}
  p_\theta(x, h) \eqdef \exp \{ \phi(x, h)^\top\theta - A(\theta) \}, \quad\text{where} \, \phi(x, h) \eqdef \phiprior(h) + \sum_{j=1}^3 \philocal(x_j, h) \in \R^d.
\end{align}

@@

\subsection{Step 2: Recover latent moments (factorization)}

@@

% Directed mixture model
\paragraph{Directed mixture model}

The directed mixture model, we have $p_\theta(x,h) = p_\theta(h) \prod_{j=1}^3 p_\theta(x_j \mid h)$,
where
$p_\theta(h) \eqdef \exp\{\phiprior(h)^\top\theta - \Aprior(\theta)\}$
and
$p_\theta(x_j \mid h) \eqdef \exp\{\philocal(x_j, h)^\top\theta - \Alocal(\theta; h)\}$.
Consequently, the joint log-likelihood is:
\begin{align}
  L(\theta) = {\underbrace{\E[\phi(x,h)]}_{\mu}}^\top\theta - \E[A(\theta; z)], \quad
  \E[A(\theta, z)] = \Aprior(\theta) + 3 \sum_{h \in [\nh]} \pi_h \Alocal(\theta;h).
\end{align}
Note that in directed models, the log-partition functions depends on the parent variables (in this case, $h$).
Having recovered the marginal distribution over $\pi$, we can compute the desired expectation.
Again, this problem is convex.

>>>>

\Fig{figures/mixtureModel}{0.3}{mixtureModel}{An (undirected) three-view
mixture model:
conditioned on a latent variable $h \in [\nh]$,
the observed variables $x_1,x_2,x_3$ are conditionally independent.}

In this section, we consider the simplest model to illustrate the form of our
method.  Consider the undirected mixture model model in \reffig{mixtureModel}.
We have one discrete latent variable $h \in [\nh]$ and
three observed variables
$x = (x_1, x_2, x_3) \in \sX_1 \times \sX_2 \times \sX_3 = \sX$
which are conditionally independent given $h$.
Formally, the model is as follows:\footnote{We assume for simplicity that the $x_j$'s are identically distributed conditioned on $h$,
but it is possible to use different distributions (see \citet{anandkumar12moments}).}
\begin{align}
  \label{eqn:mixtureModel}
  p_\theta(x, h) \eqdef \exp \{ \phi(x, h)^\top\theta - A(\theta) \}, \quad\text{where} \, \phi(x, h) \eqdef \phiprior(h) + \sum_{j=1}^3 \philocal(x_j, h) \in \R^d.
\end{align}

@@

\subsection{Step 2: Recover latent moments (factorization)}
\label{sec:factorization}

@@

### 2013-05-31 doc/nips2013/setup.tex

Let $z$ be a collection of variables indexed by $V$ (that is, $z = \{z_j\}_{j \in V}$).
Suppose that some of the variables are observed ($x = \{ z_j \}_{j \in O}$, where $O \subset V$) 
and the rest are latent ($h = \{ z_j \}_{j \in H}$, where $H = V \backslash O$).
Let $C \subset 2^V$ be a collection of subset of variables,
and let $\phi_c(z_c) \in \R^d$ be the feature map for each $c \in C$.
In this paper, we will consider both directed and undirected models over $z$.
In the undirected case, we define the standard globally-normalized exponential family model:
\begin{align}
  p_\theta(z) = \exp\{ \phi(z)^\top\theta - A(\theta) \}, \aside{undirected}
\end{align}
where $\phi(z) \eqdef \sum_{c \in C} \phi_c(z_c) \in \R^d$ is the global feature vector, $\theta \in \R^d$ is the parameter vector,
and $A(\theta) = \log \int \exp\{\phi(z)^\top\theta\} dz$ is the global log-partition function.

In the directed case, we assume a directed acyclic graph over $V$,
and define $\pi(j) \subset V$ to be the parents of each node $j \in V$.
Next, define a model as a product of locally-normalized models:
\begin{align}
  p_\theta(z) = \prod_{j \in V} p_\theta(z_j \mid z_{\pi(j)}), \quad\quad
  p_\theta(z_j \mid z_{\pi(j)}) = \exp \{ \phi_j(z_j, z_{\pi(j)})^\top\theta - A_j(\theta; z_{\pi(j)}) \}.
\end{align}
Here, each local model has its own local features $\phi_j(z_j, z_{\pi(j)})$
and its own local log-partition function $A(\theta; z_{\pi(j)})$, which depends on values of its parent.
To consolidate notation,
define $\phi(z) = \sum_{j \in V} \phi(z_j, z_{\pi(j)})$
and $A(\theta; z) = \sum_{j \in V} A_j(\theta; z_{\pi(j)})$,
so that:
\begin{align}
  p_\theta(z) \eqdef \exp\{ \phi(z)^\top\theta - A(\theta; z) \}. \aside{directed}
\end{align}

%To some extent, our methods only depend on the conditional independence
%structure of the models, not on whether the model is directed or undirected.
%Later, we will show how to estimate

%\paragraph{Example.}
%For example, \reffig{simpleModels} Markov random fields are an instance.

\paragraph{Problem statement.}

@@

%Since both models have the same sufficient statistics lead to convex
%likelihoods (assuming all variables are observed), in many cases our methods
%will not be particular to which.  In some cases, however (e.g., the factorial models),
%we will rely on marginalization properties of the directed models.

% Principles
%By insisting on developing a consistent estimator and considering identifiability,
%The key is that a variable in the model must correspond to a bottleneck.

>>>>

Let $z$ be a collection of variables indexed by $V$ (that is, $z = \{z_j\}_{j \in V}$).
Suppose that some of the variables are observed ($x = \{ z_j \}_{j \in O}$, where $O \subset V$) 
and the rest are latent ($h = \{ z_j \}_{j \in H}$, where $H = V \backslash O$).
Let $C \subset 2^V$ be a collection of subset of variables,
and let $\phi_c(z_c) \in \R^d$ be the feature map for each $c \in C$.
In this paper, we will consider both directed and undirected models over $z$.
In the undirected case, we define the standard globally-normalized exponential family model:
\begin{align}
  p_\theta(z) = \exp\{ \phi(z)^\top\theta - A(\theta) \}, \aside{undirected}
\end{align}
where $\phi(z) \eqdef \sum_{c \in C} \phi_c(z_c) \in \R^d$ is the global feature vector, $\theta \in \R^d$ is the parameter vector,
and $A(\theta) = \log \int \exp\{\phi(z)^\top\theta\} dz$ is the global log-partition function.
To some extent, our methods only depend on the conditional independence
structure of the models.
We focus on undirected models for notational simplicity;
directed models are discussed in \refapp{directedModels}.

\paragraph{Problem statement}

@@
